---
title: "Data overview"
output: rmarkdown::html_vignette
description: >
  An overview of the tables and data fields
  in mpathsenser databases.
vignette: >
  %\VignetteIndexEntry{Data overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(knitr)
library(kableExtra)
library(tibble)
library(dplyr)

table_style <- function(kbl) {
  kbl |>
    kable_styling(
      bootstrap_options = c("striped", "bordered"),
      full_width = TRUE,
      fixed_thead = TRUE
    ) |>
    row_spec(0, color = "white", background = "#0071ff") |>
    column_spec(1, extra_css = "font-family: 'Courier New';")
}

# Parse the database schema definition into a list of column data frames.
# The tables below are generated from this file, so they always match the
# schema that create_db() produces.
source_dir <- dirname(knitr::current_input(dir = TRUE))
dbdef_candidates <- c(
  file.path(source_dir, "..", "..", "inst", "extdata", "dbdef.sql"),
  file.path("inst", "extdata", "dbdef.sql"),
  system.file("extdata", "dbdef.sql", package = "mpathsenser")
)
dbdef_path <- dbdef_candidates[file.exists(dbdef_candidates)][1]
if (is.na(dbdef_path)) {
  stop("Could not find the database schema definition.")
}
dbdef <- readLines(dbdef_path)
tables <- list()
for (line in dbdef) {
  m <- regmatches(line, regexec("CREATE TABLE IF NOT EXISTS ([A-Za-z_][A-Za-z_0-9]*)\\s*\\(", line))[[1]]
  if (length(m) > 0) {
    current <- m[2]
    tables[[current]] <- data.frame(
      column = character(0), type = character(0), not_null = logical(0),
      stringsAsFactors = FALSE
    )
  } else if (exists("current") && !is.null(current)) {
    m2 <- regmatches(line, regexec("^  ([a-z_0-9]+) ([A-Za-z0-9_ ]+)( NOT NULL)?,?$", line))[[1]]
    if (length(m2) > 0) {
      tables[[current]] <- rbind(tables[[current]], data.frame(
        column = m2[2],
        type = trimws(m2[3]),
        not_null = nzchar(m2[4]),
        stringsAsFactors = FALSE
      ))
    } else if (grepl("^\\);$", line)) {
      current <- NULL
    }
  }
}

# DuckDB types to readable R-ish types
type_label <- function(type) {
  dplyr::case_when(
    type == "TEXT" ~ "character",
    type == "TIMESTAMPTZ" ~ "POSIXct (UTC)",
    type %in% c("BIGINT", "UBIGINT", "UINTEGER", "INTEGER") ~ "integer",
    type %in% c("USMALLINT", "UTINYINT", "SMALLINT") ~ "integer (small)",
    type %in% c("REAL", "DOUBLE") ~ "double",
    type == "BOOLEAN" ~ "logical",
    type == "JSON" ~ "JSON",
    TRUE ~ type
  )
}

# Column descriptions. Generic entries apply to all tables; entries prefixed
# with "Table." override the generic description for that table.
col_desc <- c(
  participant_id = "Unique ID of the participant, stored as an unsigned integer.",
  time = "Start time of the measurement, stored as a UTC instant.",
  end_time = "End time of the measurement, stored as a UTC instant.",
  source_file_id = "ID of the source file the measurement came from (see ProcessedFiles).",
  mac_address = "MAC address of the device that produced the measurement.",
  status = "Status of the measurement.",
  instance = "Identifier of the actigraphy instance.",
  bpm = "Heart rate in beats per minute.",
  bbi = "Beat-to-beat interval in milliseconds.",
  gap_duration = "Duration of the gap preceding the measurement, in milliseconds.",
  spo2 = "Blood oxygen saturation (SpO2) as a percentage.",
  temperature = "Temperature in degrees Celsius.",
  stress = "Stress score.",
  step_count = "Number of steps taken in the interval.",
  total_steps = "Total number of steps taken.",
  total_energy = "Total energy of the signal in the interval.",
  zero_crossing_count = "Number of zero crossings of the signal in the interval.",
  time_above_threshold = "Fraction of the interval during which the signal was above the threshold.",
  deadband = "Deadband value used by the sensor.",
  battery_level = "Battery level as a percentage.",
  battery_status = "Battery status (e.g. charging or discharging).",
  confidence = "Confidence of the activity recognition.",
  type = "Type of the measurement (e.g. activity type).",
  ssid = "Name of the Wi-Fi network.",
  bssid = "MAC address of the Wi-Fi access point.",
  ip = "IP address of the device on the network.",
  latitude = "Latitude in degrees.",
  longitude = "Longitude in degrees.",
  altitude = "Altitude in metres above sea level.",
  accuracy = "Accuracy of the measurement in metres.",
  vertical_accuracy = "Vertical accuracy of the measurement in metres.",
  speed = "Speed in metres per second.",
  speed_accuracy = "Accuracy of the speed measurement in metres per second.",
  heading = "Heading in degrees.",
  heading_accuracy = "Accuracy of the heading measurement in degrees.",
  is_mock = "Whether the measurement was produced by a mock location provider.",
  elapsed_realtime_nanos = "Elapsed realtime since boot at the time of the measurement, in nanoseconds.",
  elapsed_realtime_uncertainty_nanos = "Uncertainty of the elapsed realtime, in nanoseconds.",
  region = "Region identifier of the beacon.",
  uuid = "UUID of the beacon.",
  major = "Major identifier of the beacon.",
  minor = "Minor identifier of the beacon.",
  proximity = "Proximity of the beacon.",
  rssi = "Received signal strength indicator in dBm.",
  tx_power_level = "Transmit power level of the beacon in dBm.",
  advertisement_name = "Bluetooth advertisement name.",
  bluetooth_device_id = "Identifier of the Bluetooth device.",
  bluetooth_device_name = "Name of the Bluetooth device.",
  connectable = "Whether the Bluetooth device is connectable.",
  connectivity_status = "Connectivity status (e.g. wifi, mobile, none).",
  device_id = "Identifier of the device.",
  hardware = "Hardware of the device.",
  device_name = "Name of the device.",
  device_manufacturer = "Manufacturer of the device.",
  device_model = "Model of the device.",
  operating_system = "Operating system of the device.",
  platform = "Platform of the device.",
  operating_system_version = "Version of the operating system.",
  device_data = "Raw device data as recorded by the sensor.",
  message = "Error message.",
  free_physical_memory = "Free physical memory in bytes.",
  free_virtual_memory = "Free virtual memory in bytes.",
  period = "Period of the heartbeat, in seconds.",
  device_type = "Type of the device.",
  device_role_name = "Role of the device.",
  mean_lux = "Mean illuminance in lux over the window.",
  std_lux = "Standard deviation of the illuminance in lux over the window.",
  min_lux = "Minimum illuminance in lux over the window.",
  max_lux = "Maximum illuminance in lux over the window.",
  country = "Country of the location.",
  area_name = "Name of the area.",
  weather_main = "Main weather condition.",
  weather_description = "Description of the weather condition.",
  sunrise = "Time of sunrise, stored as a UTC instant.",
  sunset = "Time of sunset, stored as a UTC instant.",
  pressure = "Atmospheric pressure in hPa.",
  wind_speed = "Wind speed in metres per second.",
  wind_degree = "Wind direction in degrees.",
  humidity = "Relative humidity as a percentage.",
  cloudiness = "Cloudiness as a percentage.",
  rain_last_hour = "Rainfall in the last hour in millimetres.",
  rain_last_3hours = "Rainfall in the last 3 hours in millimetres.",
  snow_last_hour = "Snowfall in the last hour in millimetres.",
  snow_last_3hours = "Snowfall in the last 3 hours in millimetres.",
  temperature_weather = "Temperature in degrees Celsius.",
  temp_min = "Minimum temperature in degrees Celsius.",
  temp_max = "Maximum temperature in degrees Celsius.",
  timezone = "Timezone of the measurement (e.g. Europe/Brussels).",
  screen_event = "Screen event (e.g. screen on or off).",
  usage = "Duration of app use in milliseconds.",
  app = "Name of the app.",
  package_name = "Package name of the app.",
  last_foreground = "Time the app was last brought to the foreground, stored as a UTC instant.",
  period_start = "Start of the usage period, stored as a UTC instant.",
  period_end = "End of the usage period, stored as a UTC instant.",
  n_accelerometer = "Number of accelerometer measurements in the interval.",
  n_actigraphy_1 = "Number of actigraphy (instance 1) measurements in the interval.",
  n_actigraphy_2 = "Number of actigraphy (instance 2) measurements in the interval.",
  n_actigraphy_3 = "Number of actigraphy (instance 3) measurements in the interval.",
  n_bbi = "Number of BBI measurements in the interval.",
  n_enhanced_bbi = "Number of enhanced BBI measurements in the interval.",
  n_gyroscope = "Number of gyroscope measurements in the interval.",
  n_heartrate = "Number of heart rate measurements in the interval.",
  n_respiration = "Number of respiration measurements in the interval.",
  n_skin_temperature = "Number of skin temperature measurements in the interval.",
  n_spo2 = "Number of SpO2 measurements in the interval.",
  n_steps = "Number of step measurements in the interval.",
  n_stress = "Number of stress measurements in the interval.",
  n_wrist_status = "Number of wrist status measurements in the interval.",
  n_zero_crossing = "Number of zero crossing measurements in the interval.",
  # GarminX/Y/Z columns and accelerometer feature columns
  "GarminAccelerometer.x" = "Acceleration along the x axis in g.",
  "GarminAccelerometer.y" = "Acceleration along the y axis in g.",
  "GarminAccelerometer.z" = "Acceleration along the z axis in g.",
  "GarminGyroscope.x" = "Angular velocity around the x axis in degrees per second.",
  "GarminGyroscope.y" = "Angular velocity around the y axis in degrees per second.",
  "GarminGyroscope.z" = "Angular velocity around the z axis in degrees per second.",
  "GarminMeta.time_from" = "Start of the logging interval, stored as a UTC instant.",
  "GarminMeta.time_to" = "End of the logging interval, stored as a UTC instant."
)

accel_axis <- c(x = "x", y = "y", z = "z")
accel_desc <- c(
  mean = "Mean value of the acceleration along the {axis} axis over the window.",
  median = "Median value of the acceleration along the {axis} axis over the window.",
  std = "Standard deviation of the acceleration along the {axis} axis over the window.",
  aad = "Average absolute deviation of the acceleration along the {axis} axis over the window.",
  min = "Minimum value of the acceleration along the {axis} axis over the window.",
  max = "Maximum value of the acceleration along the {axis} axis over the window.",
  max_min_diff = "Difference between the maximum and minimum acceleration along the {axis} axis over the window.",
  mad = "Median absolute deviation of the acceleration along the {axis} axis over the window.",
  iqr = "Interquartile range of the acceleration along the {axis} axis over the window.",
  neg_n = "Number of negative acceleration values along the {axis} axis over the window.",
  pos_n = "Number of positive acceleration values along the {axis} axis over the window.",
  above_mean = "Number of acceleration values above the mean along the {axis} axis over the window.",
  energy = "Energy of the acceleration signal along the {axis} axis over the window."
)

describe <- function(table, column) {
  key <- paste0(table, ".", column)
  if (key %in% names(col_desc)) return(col_desc[[key]])
  if (column %in% names(col_desc)) return(col_desc[[column]])
  if (table == "Accelerometer" && grepl("^([xyz])_([a-z_]+)$", column)) {
    axis <- sub("^([xyz])_.*$", "\\1", column)
    stat <- sub("^[xyz]_(.*)$", "\\1", column)
    if (stat %in% names(accel_desc)) {
      return(gsub("\\{axis\\}", accel_axis[[axis]], accel_desc[[stat]], fixed = TRUE))
    }
  }
  "Value of the field as recorded by the sensor."
}

# Render the schema table of one sensor table as a styled kable
schema_table <- function(table) {
  df <- tables[[table]]
  df |>
    transmute(
      Variable = column,
      Type = type_label(type),
      Description = vapply(column, function(c) describe(table, c), character(1)),
      `Not NULL` = ifelse(not_null, "yes", "")
    ) |>
    mutate(Variable = cell_spec(Variable, underline = Variable %in% c("participant_id", "time"))) |>
    kbl(booktabs = TRUE, escape = FALSE) |>
    table_style()
}
```

This article provides an overview of the tables in an `mpathsenser` database. The database is created by `create_db()` and populated by `read_mpath_sense()`. It is a [DuckDB](https://duckdb.org) database, so all data is stored in a columnar format and queried through SQL (via `dplyr`/`dbplyr` or `DBI`). All timestamps are stored as UTC instants.

The database consists of two parts: metadata tables (`Study`, `Participant`, `ProcessedFiles`, and `Meta`) and one table per sensor. Every sensor table contains the `participant_id` of the participant who produced the measurement and the `source_file_id` of the file it came from, which links each measurement to its provenance in `ProcessedFiles`. The `participant_id` is stored as an unsigned integer, matching how the m-Path Sense `connectionId` is encoded. The tables below are generated from the schema definition (`dbdef.sql`) that `create_db()` uses, so they always match the initial database schema.

`ProcessedFiles` identifies an imported file by the combination of `file_name`, `participant_id`, `file_size_bytes`, and `modified_at`. There is deliberately no content-hash column. The `file_id` is also used as `source_file_id` in sensor tables and determines which file wins when duplicate measurements are deduplicated.

The initial sensor tables have no unique constraints: duplicate measurements are possible (e.g. the same file imported under a different name) and are removed by the deduplication step of `read_mpath_sense()`. Per measurement key (`participant_id`, `time`, plus sensor-specific extras such as the app for `AppUsage`), the row of the newest file is kept. Within that newest file, most sensors keep the first row; interval sensors (those with an `end_time`) and Garmin's recalculated point measurements (`GarminBBI`, `GarminEnhancedBBI`, `GarminHeartRate`, `GarminStress`) keep the last row, since they can legitimately repeat a start time with a later `end_time` or a recalculated value. See the deduplication section of the main workflow vignette for details.

## Querying the database

`get_data()` returns a lazy `dbplyr` table. Filter and summarise data before calling `collect()` so that DuckDB performs the work rather than copying the whole sensor table into R:

```{r, eval=FALSE}
library(dplyr)

get_data(db, sensor = "Battery") |>
  filter(participant_id == "12345") |>
  summarise(mean_battery = mean(battery_level, na.rm = TRUE)) |>
  collect()
```

Because the database is DuckDB, it can also be inspected directly through DBI:

```{r, eval=FALSE}
DBI::dbListTables(db)
DBI::dbGetQuery(
  db,
  "SELECT participant_id, COUNT(*) AS n
   FROM Battery
   GROUP BY participant_id
   ORDER BY participant_id"
)
```

For dashboards or other readers, open a separate read-only connection with `open_db(..., read_only = TRUE)`. DuckDB supports multiple readers, but simultaneous imports or other writers are not supported. Set `temp_directory` to a local, non-synchronised directory when queries may spill to disk; this is especially important when the database file is stored in OneDrive.

## Meta-data

The schema tables below describe the columns created by `create_db()`. Sensor tables contain both `TIMESTAMPTZ` measurement timestamps and a `timezone` column. The latter is populated automatically during import when `Timezone` measurements are available.

### Meta

The `Meta` table stores key-value pairs for database metadata.

```{r meta-table, echo=FALSE}
schema_table("Meta")
```

### Study

This table contains information about the study itself. The data from this table can be accessed using `get_studies()`.

```{r study-table, echo=FALSE}
schema_table("Study")
```

### Participant

This table contains information about the participants. The data from this table can be accessed using `get_participants()`.

```{r participant-table, echo=FALSE}
schema_table("Participant")
```

### ProcessedFiles

This table keeps track of which files have already been imported, so that `read_mpath_sense()` only processes new files on subsequent runs. A file is considered processed when a record with the same file name, participant, size, and modification time exists. The data from this table can be accessed using `get_processed_files()`. The `file_id` is used as `source_file_id` in the sensor tables and decides which file wins on deduplication.

```{r processedfiles-table, echo=FALSE}
schema_table("ProcessedFiles")
```

## Sensor-specific tables

The tables below describe the physical sensor tables in the `main` schema. Canonical timestamps are stored as UTC `TIMESTAMPTZ` values, and each row may include an IANA `timezone` for local rendering. Explicitly named local views expose participant-local wall-clock values; `COALESCE(timezone, 'UTC')` ensures unmapped observations still have non-NULL timestamps. The import populates timezone values automatically; `add_timezones_to_db()` can be rerun to normalize existing or selected tables.

```{r, echo=FALSE, results="asis"}
metadata_tables <- c("Meta", "Study", "Participant", "ProcessedFiles")
schema_sensors <- setdiff(names(tables), metadata_tables)
non_garmin <- setdiff(schema_sensors, grep("^Garmin", schema_sensors, value = TRUE))
lines <- unlist(lapply(non_garmin, function(s) {
  c(
    paste0("### ", s, "\n"),
    paste0("```{r ", tolower(s), "-table, echo=FALSE}"),
    paste0('schema_table("', s, '")'),
    "```",
    ""
  )
}))
cat(knitr::knit_child(text = lines, quiet = TRUE))
```

## Garmin sensor tables

The tables below describe the data collected from Garmin devices paired with the phone (direct sensing). These sensors are collected in batches: each `garminalllogsdata` entry in the raw data can contain thousands of measurements, which are stored as one row per measurement. The `GarminMeta` table summarises per participant how many measurements of each type were collected in a logging interval.

```{r, echo=FALSE, results="asis"}
garmin <- grep("^Garmin", schema_sensors, value = TRUE)
lines <- unlist(lapply(garmin, function(s) {
  c(
    paste0("### ", s, "\n"),
    paste0("```{r ", tolower(s), "-table, echo=FALSE}"),
    paste0('schema_table("', s, '")'),
    "```",
    ""
  )
}))
cat(knitr::knit_child(text = lines, quiet = TRUE))
```
